A Dependently Typed Multi-Stage Calculus

### 2019-08-28 text/related-work.tex
Add some comments.
Reflecting taha's comment.

% 多段階計算の歴史

Theoretical studies on multi-stage programming owe a lot to seminal
work by Davies and Pfenning~\cite{DaviesPfenning01JACM} and
Davies~\cite{davies1996temporal}, who found Curry-Howard
correspondence between multi-stage calculi and modal logic.  In
particular, Davies' $\lambda^\circ$~\cite{davies1996temporal} has been
a basis for several multi-stage calculi with quasi-quotation.  $\lambda^\circ$
did not have operators for run and CSP; a few
studies~\cite{benaissa1999logical,MoggiTahaBenaissaSheard99ESOP}
enhanced and improved $\lambda^\circ$ towards the development of a
type-safe multi-stage calculus with quasi-quotation, run, and CSP,
which were proposed by Taha and Sheard as constructs for multi-stage
programming~\cite{MetaML}.
% Benaissa et al.~\cite{benaissa1999logical} study the relationship between a multi-stage calculus and category theory or modal logic.
% Taha and Sheard introduced run and CSP to a multi-stage calculus in 
Finally, Taha and Nielsen invented the concept of environment
classifiers~\cite{taha2003environment} and developed a typed calculus
$\lambda^\alpha$, which was equipped with all the features above in a
type sound manner and formed a basis of earlier versions of MetaOCaml.
Different approaches to type-safe multi-stage programming with
slightly different constructs for composing and running code values have been studied by Kim, Yi,
and Calcagno~\cite{DBLP:conf/popl/KimYC06} and Nanevski and Pfenning~\cite{DBLP:journals/jfp/NanevskiP05}.

@@

% 他段階計算の機能拡張

% 多段階計算の応用

% 依存型の歴史

@@

% 依存型の応用

% Practical applications of dependent types have been also studied.
% One can use dependent types in programming languages such as Idris~\cite{brady2013idris} or
% interactive theorem provers such as Coq~\cite{09thecoq} based on \cite{coquand:inria-00076024}.
% In Xi and Pfenning~\cite{Xi98}, they extended SML with restricted dependent types
% and succeeded in reducing the bounds checking of arrays.
% In Xi and Harper~\cite{xi2001dependently}, they design a type system for an assembly language and
% it is useful for speed up.
% Xi also gave dead code elimination and loop unrolling as applications
%  of dependent types~\cite{xi1999dependent}.

% 関連研究との比較

Although there are studies on combinations of multi-stage programming and
other programming features such as mutable cells~\cite{kiselyov2016refined} and
control operators~\cite{KameyamaKiselyovShan09PEPM,oishi2017staging}, a
combination with dependent types has been little studied. One exception is
Brady and Hammond~\cite{brady2006dependently}, who have discussed a
combination of dependently typed programming with staging in the style of
MetaOCaml to implement a staged interpreter, which is statically guaranteed
to generate well-typed code. However, they focused on concrete programming
examples and there is no theoretical investigation of the programming
language they used. In order to make safer multi-stage programming language, 
Martin and Laurence~\cite{martin2015HGRTMP} gave program logic for
\(\text{Mini-ML}^\square_e\)~\cite{DaviesPfenning01JACM} but
it cannot manipulate open code which ours can deal with.

@@

% Although we define type equivalence of \LMD with composition of equivalence rules,
% there is another candidate to define,
% which gives reduction on types and compare the results of reduction such as~\cite{sorensen2006lectures}.
% This method is better than one of \LMD because equivalence rules become simple.
% However, we reject it because it cannot handle CSP flexibly.

>>>>

% The history of multi-stage programming

Theoretical studies on multi-stage programming owe a lot to seminal work by
Davies and Pfenning~\cite{DaviesPfenning01JACM} and
Davies~\cite{davies1996temporal}, who found Curry-Howard correspondence
between multi-stage calculi and modal logic. In particular, Davies'
$\lambda^\circ$~\cite{davies1996temporal} has been a basis for several
multi-stage calculi with quasi-quotation. $\lambda^\circ$ did not have
operators for run and CSP; a few
studies~\cite{benaissa1999logical,MoggiTahaBenaissaSheard99ESOP} enhanced and
improved $\lambda^\circ$ towards the development of a type-safe multi-stage
calculus with quasi-quotation, run, and CSP, which were proposed by Taha and
Sheard as constructs for multi-stage programming~\cite{MetaML}.

Finally, Taha and Nielsen invented the concept of environment
classifiers~\cite{taha2003environment} and developed a typed calculus
$\lambda^\alpha$, which was equipped with all the features above in a type
sound manner and formed a basis of earlier versions of MetaOCaml. Different
approaches to type-safe multi-stage programming with slightly different
constructs for composing and running code values have been studied by Kim,
Yi, and Calcagno~\cite{DBLP:conf/popl/KimYC06} and Nanevski and
Pfenning~\cite{DBLP:journals/jfp/NanevskiP05}.

@@

% The history of dependent types

@@

% Applications of dependent types

% Practical applications of dependent types have been also studied. One can use
% dependent types in programming languages such as Idris~\cite{brady2013idris}
% or interactive theorem provers such as Coq~\cite{09thecoq} based on
% \cite{coquand:inria-00076024}. In Xi and Pfenning~\cite{Xi98}, they extended
% SML with restricted dependent types and succeeded in reducing the bounds
% checking of arrays. In Xi and Harper~\cite{xi2001dependently}, they design a
% type system for an assembly language and it is useful for speed up. Xi also
% gave dead code elimination and loop unrolling as applications of dependent
% types~\cite{xi1999dependent}.

% Comparison with other works

Although there are studies on combinations of multi-stage programming and
other programming features such as mutable cells~\cite{kiselyov2016refined} and
control operators~\cite{KameyamaKiselyovShan09PEPM,oishi2017staging}, a
combination with dependent types has been little studied.

Brady and Hammond~\cite{brady2006dependently} have discussed a
combination of dependently typed programming with staging in the style of
MetaOCaml to implement a staged interpreter, which is statically guaranteed
to generate well-typed code. However, they focused on concrete programming
examples and there is no theoretical investigation of the programming
language they used.

Pasalic and Taha~\cite{pasalic2002tagless} designed \(\lambda_{H\circ}\) by
introducing the concept of stage into an existing dependent type system
\(\lambda_H\)~\cite{zhong2002certified}. However, there are not eval and CSP
in \(\lambda_{H\circ}\) because it was designed before the invention of
environment classifiers.

In Forgarty, Pasalic, Siek and Taha~\cite{fogarty2007concoqtion}

In order to make safer multi-stage programming language, 
Martin and Laurence~\cite{martin2015HGRTMP} gave program logic for
\(\text{Mini-ML}^\square_e\)~\cite{DaviesPfenning01JACM} but
it cannot manipulate open code which ours can deal with. \AK{Is the place of
this sentence is appropriate?}

@@

Although we define type equivalence of \LMD with composition of equivalence
rules, there is another candidate to define, which gives reduction on types
and compare the results of reduction such as~\cite{sorensen2006lectures}.
This method is better than one of \LMD because equivalence rules become
simple. However, we reject it because it cannot handle CSP flexibly.
